```{r setup}
#| message: false
#| include: false
library(tidyverse)
library(knitr)
```

# Raw {.unnumbered}

First we explored raw species data content. This corresponds to **version 2**.

## Taxonomy preparation

Taxonomic information for Moju and Peteco was only contained in the vernacular column, while similarly taxonomic information for Kibale, Lesong, Mbaiki, Misiones, Ulu Muda, Sg Lalang was only contained in the scientific column. We thus smoothed all taxonomic information with five columns: scientific, vernacular, family, genus, species (`derived_data/raw_taxo_preped_v2.tsv`).

```{r taxo_prep}
#| message: false
#| warning: false
taxo_easy <- read_tsv("data/raw_data/raw_taxonomy_v1.tsv") %>%
  rename_all(tolower) %>%
  filter(!(site %in% c(
    "Kibale", "Lesong", "Moju", "Mbaiki", "Peteco",
    "Misiones", "Ulu Muda", "Sg Lalang"
  )))
taxo_vernacular_only <- read_tsv("data/raw_data/raw_taxonomy_v1.tsv") %>%
  rename_all(tolower) %>%
  filter(site %in% c("Moju", "Peteco")) %>%
  separate(vernname, c("vernname", "scientificname"), sep = "\\[") %>%
  separate(vernname, "vernname", sep = "\\/") %>%
  mutate(scientificname = gsub("]", "", scientificname)) %>%
  separate(scientificname, c("genus", "species"), remove = FALSE)
taxo_scientific_only <- read_tsv("data/raw_data/raw_taxonomy_v1.tsv") %>%
  rename_all(tolower) %>%
  filter(site %in% c(
    "Kibale", "Lesong", "Mbaiki",
    "Misiones", "Ulu Muda", "Sg Lalang"
  )) %>%
  separate(scientificname, c("genus", "species"), remove = FALSE)
taxo <- bind_rows(taxo_easy, taxo_vernacular_only, taxo_scientific_only) %>%
  rename(scientific = scientificname, vernacular = vernname)
write_tsv(taxo, "data/derived_data/raw_taxo_preped_v2.tsv")
```

## All

We had data for 16 sites with 0 to 68 families, with 49 to 301 genera, and with 123 to 762 species corresponding to 0 to 364 vernacular names when registered:

```{r stats}
taxo %>%
  group_by(site) %>%
  summarise(
    family = length(unique(na.omit(family))),
    genus = length(unique(na.omit(genus))),
    species = length(unique(paste(genus, species))),
    scientific = length(unique(na.omit(scientific))),
    vernacular = length(unique(na.omit(vernacular)))
  ) %>%
  kable(caption = "Numbers per taxonomic levels for each site.")
```

In total, we thus add 1,611 species before cleaning:

```{r totals}
taxo %>%
  summarise(
    family = length(unique(na.omit(family))),
    genus = length(unique(na.omit(genus))),
    species = length(unique(paste(genus, species))),
    scientific = length(unique(na.omit(scientific))),
    vernacular = length(unique(na.omit(vernacular)))
  ) %>%
  kable(
    format.args = list(big.mark = ","),
    caption = "Total numbers per taxonomic levels."
  )
```

Species name construction using genus and species should be preferred as the scientific name column increases diversity with only slight differences:

```{r diffs}
taxo %>%
  mutate(species = paste(genus, species)) %>%
  select(species, scientific) %>%
  unique() %>%
  group_by(species) %>%
  mutate(N = n()) %>%
  ungroup() %>%
  filter(N == 12) %>%
  kable(caption = "Example of multiple scientific names for one species genus pair.") # nolint
```

```{r diffs_fig}
#| message: false
#| fig-cap: Scientific names per genus species pairs.
taxo %>%
  mutate(species = paste(genus, species)) %>%
  select(species, scientific) %>%
  unique() %>%
  group_by(species) %>%
  summarise(N = n()) %>%
  filter(N > 1) %>%
  ggplot(aes(N)) +
  geom_histogram() +
  theme_bw() +
  scale_y_sqrt() +
  scale_x_sqrt() +
  xlab("Scientific names per genus species pairs") +
  ylab("")
```

## Family

Families includes the following dubious names:

-   **Missing**: #N/A, +, muerto
-   **Unidentified**: Indet., Not in DB, unid., unidentified, ND, No determinado

See the list below:

```{r family}
unique(taxo$family) %>%
  sort() %>%
  paste0(collapse = ", ")
```

With the following counts:

```{r familytab}
unid <- c(
  "Indet.", "Not in DB", "unid.",
  "unidentified", "ND", "No determinado"
)
missing <- c("#N/A", "+", "muerto")
taxo %>%
  select(family) %>%
  unique() %>%
  summarise(
    "Unidentified" = sum(family %in% unid,
      na.omit = TRUE
    ),
    "Missing" = sum(family %in% missing,
      na.omit = TRUE
    )
  ) %>%
  gather("type", "number") %>%
  kable(caption = "Number of family by issue type.")
```

## Genus

Genera includes the following dubious names:

-   **Missing**: #N/A, A
-   **Unidentified**: Indet., nd, sp, sp1, sp5, Indetermined, KERUING, MERANTI, MERSAWA, BALAU, muerto
-   **Family-only**: Indet.Apocynaceae, Indet.Arecaceae, Indet.Bignoniaceae, Indet.Burseraceae, Indet.Capparaceae, Indet.Chrysobalanaceae, Indet.Clusiaceae, Indet.Combretaceae, Indet.Dichapetalaceae, Indet.Euphorbiaceae, Indet.Fabaceae, Indet.FabaceaeCaesalpinioideae(mimosoid_clade), Indet.FabaceaePapilionoideae, Indet.Humiriaceae, Indet.Indet., Indet.Lamiaceae, Indet.Lauraceae, Indet.Lecythidaceae, Indet.Malvaceae, Indet.MalvaceaeBombacoideae, Indet.Melastomataceae, Indet.Moraceae, Indet.Myrtaceae, Indet.Nyctaginaceae, Indet.Olacaceae, Indet.Rubiaceae, Indet.Salicaceae, Indet.Sapindaceae, Indet.Sapotaceae, Indet.Urticaceae, Indet.Violaceae, Indet.Vochysiaceae, Indet.Anacardiaceae, Indet.Annonaceae

See the list below:

```{r genus}
unique(taxo$genus) %>%
  sort() %>%
  paste0(collapse = ", ")
```

With the following counts:

```{r genustab}
taxo %>%
  select(genus) %>%
  unique() %>%
  summarise(
    "Missing" = sum((genus %in% c("#N/A", "A")),
      na.omit = TRUE
    ),
    "Indet" = sum(
      (genus %in% c(
        "Indet.", "nd", "sp", "sp1", "sp5", "Indetermined",
        "No", "KERUING", "MERANTI", "MERSAWA",
        "BALAU", "muerto"
      )),
      na.omit = TRUE
    ),
    "Family-only" = sum(grepl("Indet.", genus), na.omit = TRUE)
  ) %>%
  gather("type", "number") %>%
  kable(caption = "Number of genus by issue type.")
```

## Species

Species includes the following dubious names:

-   **Unidentified** ("det", "\^sp"). Indet., sp, sp., Schaeff., Mast, Welw, Stapf, Hook, Gilg, Benth, flacouirtiaceae, Determinado, BERMINYAK, TIDAK,KUNING, spp,
-   sp1, sp5, L, muerto
-   **Sub-taxonomic** ("\[\[:punct:\]\]"): guyanensis subsp. Guyanensis, acuminatum_subsp.juniperinum, cuspidata_subsp.dura, guyanensis subsp. Guyanensis, guyanensis_subsp.duckeana, opacum_subsp.rabelianum, panacoco_var.panacoco, persistens_subsp.aurantiaca, tomentosa_subsp.maroniensis, venosa_subsp.amazonica, vera_subsp.affinis
-   **Undescribed** ("\^sp"). sp.1, sp.1-CAY, sp.10-CAY, sp.11-CAY, sp.12-CAY, sp.13-CAY, sp.14-CAY, sp.15-CAY, sp.16-CAY, sp.17-CAY, sp.18-CAY, sp.1Guyafor, sp.2, sp.2-CAY, sp.21_DS, sp.22-CAY, sp.23-CAY, sp.24-CAY, sp.2BAF, sp.2Guyafor, sp.3, sp.3-CAY, sp.30-CAY, sp.33-CAY, sp.34-CAY, sp.35-CAY, sp.35Guyafor, sp.36-CAY, sp.38Guyafor, sp.3BAF, sp.3Guyafor, sp.4-CAY, sp.42-CAY, sp.5-CAY, sp.5Guyafor, sp.6-CAY, sp.6Guyafor, sp.7-CAY, sp.70-CAY, sp.71-CAY, sp.7Guyafor, sp.8-CAY, sp.9-CAY, sp.B1, sp.B10, sp.B2, sp.B35, sp.B5, sp.B8, sp.B9, sp.C, sp.D, sp.FG13-Holst, sp.FG14-Holst, sp.FG2-Holst, sp.FG21-Holst, sp.FG9-Holst, sp.P1, sp.P11, sp.P22, sp.P3, sp.P33, sp.P4, sp.P6, sp.P9

See the list below:

```{r species}
t <- select(taxo, species) %>%
  unique() %>%
  filter(grepl("^sp", species))
unique(taxo$species) %>%
  sort() %>%
  paste0(collapse = ", ")
```

```{r speciestab}
unid <- c(
  "Indet.", "sp", "sp.", "Schaeff.",
  "Mast", "Welw", "Stapf", "Hook", "Gilg", "Benth",
  "flacouirtiaceae", "Determinado", "BERMINYAK", "TIDAK", "KUNING", "spp",
  "sp1", "sp5", "L", "muerto"
)
subtaxo <- c(
  "guyanensis subsp. Guyanensis", "acuminatum_subsp.juniperinum",
  "cuspidata_subsp.dura", "guyanensis subsp. Guyanensis",
  "guyanensis_subsp.duckeana", "opacum_subsp.rabelianum",
  "panacoco_var.panacoco", "persistens_subsp.aurantiaca",
  "tomentosa_subsp.maroniensis", "venosa_subsp.amazonica",
  "vera_subsp.affinis"
)
undesc <- c(
  "sp.1", "sp.1-CAY", "sp.10-CAY", "sp.11-CAY", "sp.12-CAY",
  "sp.13-CAY", "sp.14-CAY", "sp.15-CAY", "sp.16-CAY",
  "sp.17-CAY", "sp.18-CAY", "sp.1Guyafor", "sp.2", "sp.2-CAY",
  "sp.21_DS", "sp.22-CAY", "sp.23-CAY", "sp.24-CAY", "sp.2BAF",
  "sp.2Guyafor", "sp.3", "sp.3-CAY", "sp.30-CAY", "sp.33-CAY",
  "sp.34-CAY", "sp.35-CAY", "sp.35Guyafor", "sp.36-CAY",
  "sp.38Guyafor", "sp.3BAF", "sp.3Guyafor", "sp.4-CAY",
  "sp.42-CAY", "sp.5-CAY", "sp.5Guyafor", "sp.6-CAY", "sp.6Guyafor",
  "sp.7-CAY", "sp.70-CAY", "sp.71-CAY", "sp.7Guyafor", "sp.8-CAY",
  "sp.9-CAY", "sp.B1", "sp.B10", "sp.B2", "sp.B35", "sp.B5", "sp.B8",
  "sp.B9", "sp.C", "sp.D", "sp.FG13-Holst", "sp.FG14-Holst",
  "sp.FG2-Holst", "sp.FG21-Holst", "sp.FG9-Holst", "sp.P1", "sp.P11",
  "sp.P22", "sp.P3", "sp.P33", "sp.P4", "sp.P6", "sp.P9"
)
taxo %>%
  select(species) %>%
  unique() %>%
  summarise(
    "Unidentified" = sum(species %in% unid,
      na.omit = TRUE
    ),
    "Sub-taxonomic" = sum(species %in% subtaxo,
      na.omit = TRUE
    ),
    "Undescribed" = sum(species %in% undesc,
      na.omit = TRUE
    )
  ) %>%
  gather("type", "number") %>%
  kable(caption = "Number of species by issue type.")
```
