```{r setup}
#| message: false
#| include: false
library(tidyverse)
library(knitr)
```

# Raw {.unnumbered}

First we explored raw species data content. This corresponds to **version 1**.

## All

We first removed Kibale, Lesong, Misiones, Mbaiki, Moju, Peteco, Sg Lalang, and Ulu Muda as taxonomic information is currently contained only either in the vernacular or scientific column.

> **Version 2 should include specific cleaning of sites with know issue as discussed with Camille.**

```{r taxo}
#| message: false
taxo <- read_tsv("data/raw_data/raw_taxonomy.tsv") %>%
  rename_all(tolower) %>%
  filter(!(site %in% c(
    "Kibale", "Lesong", "Moju", "Mbaiki", "Peteco",
    "Misiones", "Ulu Muda", "Sg Lalang"
  )))
```

We had data for 8 sites:

```{r sites}
taxo %>%
  select(site) %>%
  unique() %>%
  summarise(sites = n()) %>%
  kable(caption = "Total number of sites.")
```

We had a 43 to 69 families, with 101 to 256 genera, and with 130 to 762 species corresponding to 198 to 247 vernacular names when registered:

```{r stats}
taxo %>%
  group_by(site) %>%
  summarise(
    family = length(unique(na.omit(family))),
    genus = length(unique(na.omit(genus))),
    species = length(unique(paste(genus, species))),
    scientific = length(unique(na.omit(scientificname))),
    vernacular = length(unique(na.omit(vernname)))
  ) %>%
  kable(caption = "Numbers per taxonomic levels for each site.")
```

In total, we thus add 1,140 species before cleaning:

```{r totals}
taxo %>%
  summarise(
    family = length(unique(na.omit(family))),
    genus = length(unique(na.omit(genus))),
    species = length(unique(paste(genus, species))),
    scientific = length(unique(na.omit(scientificname))),
    vernacular = length(unique(na.omit(vernname)))
  ) %>%
  kable(
    format.args = list(big.mark = ","),
    caption = "Total numbers per taxonomic levels."
  )
```

Species name construction using genus and species should be preferred as the scientific name column increases diversity with only slight differences:

```{r diffs}
taxo %>%
  mutate(species = paste(genus, species)) %>%
  select(species, scientificname) %>%
  unique() %>%
  group_by(species) %>%
  mutate(N = n()) %>%
  ungroup() %>%
  filter(N == 11) %>%
  kable(caption = "Example of multiple scientific names for one species genus pair.") # nolint
```

```{r diffs_fig}
#| message: false
#| fig-cap: Scientific names per genus species pairs.
taxo %>%
  mutate(species = paste(genus, species)) %>%
  select(species, scientificname) %>%
  unique() %>%
  group_by(species) %>%
  summarise(N = n()) %>%
  filter(N > 1) %>%
  ggplot(aes(N)) +
  geom_histogram() +
  theme_bw() +
  scale_y_sqrt() +
  scale_x_sqrt() +
  xlab("Scientific names per genus species pairs") +
  ylab("")
```

## Family

Families includes the following dubious names:

-   **Missing**: #N/A, +
-   **Unidentified**: Indet., Not in DB, unid., unidentified

See the list below:

```{r family}
unique(taxo$family) %>%
  sort() %>%
  paste0(collapse = ", ")
```

With the following counts:

```{r familytab}
unid <- c("Indet.", "Not in DB", "unid.", "unidentified")
missing <- c("#N/A", "+")
taxo %>%
  select(family) %>%
  unique() %>%
  summarise(
    "Unidentified" = sum(family %in% unid,
      na.omit = TRUE
    ),
    "Missing" = sum(family %in% missing,
      na.omit = TRUE
    )
  ) %>%
  gather("type", "number") %>%
  kable(caption = "Number of family by issue type.")
```

## Genus

Genera includes the following dubious names:

-   **Missing**: #N/A,
-   **Unidentified**: Indet.
-   **Family-only**: Indet.Apocynaceae, Indet.Arecaceae, Indet.Bignoniaceae, Indet.Burseraceae, Indet.Capparaceae, Indet.Chrysobalanaceae, Indet.Clusiaceae, Indet.Combretaceae, Indet.Dichapetalaceae, Indet.Euphorbiaceae, Indet.Fabaceae, Indet.FabaceaeCaesalpinioideae(mimosoid_clade), Indet.FabaceaePapilionoideae, Indet.Humiriaceae, Indet.Indet., Indet.Lamiaceae, Indet.Lauraceae, Indet.Lecythidaceae, Indet.Malvaceae, Indet.MalvaceaeBombacoideae, Indet.Melastomataceae, Indet.Moraceae, Indet.Myrtaceae, Indet.Nyctaginaceae, Indet.Olacaceae, Indet.Rubiaceae, Indet.Salicaceae, Indet.Sapindaceae, Indet.Sapotaceae, Indet.Urticaceae, Indet.Violaceae, Indet.Vochysiaceae

See the list below:

```{r genus}
unique(taxo$genus) %>%
  sort() %>%
  paste0(collapse = ", ")
```

With the following counts:

```{r genustab}
taxo %>%
  select(genus) %>%
  unique() %>%
  summarise(
    "Missing" = sum((genus %in% c("#N/A", "Indet.")),
      na.omit = TRUE
    ),
    "Family-only" = sum(grepl("Indet.", genus), na.omit = TRUE)
  ) %>%
  gather("type", "number") %>%
  kable(caption = "Number of genus by issue type.")
```

## Species

Species includes the following dubious names:

-   **Unidentified** ("det", "\^sp"). Indet., sp, sp., Schaeff.
-   **Sub-taxonomic** ("\[\[:punct:\]\]"): guyanensis subsp. Guyanensis, acuminatum_subsp.juniperinum, cuspidata_subsp.dura, guyanensis subsp. Guyanensis, guyanensis_subsp.duckeana, opacum_subsp.rabelianum, panacoco_var.panacoco, persistens_subsp.aurantiaca, tomentosa_subsp.maroniensis, venosa_subsp.amazonica, vera_subsp.affinis
-   **Undescribed** ("\^sp"). sp.1, sp.1-CAY, sp.10-CAY, sp.11-CAY, sp.12-CAY, sp.13-CAY, sp.14-CAY, sp.15-CAY, sp.16-CAY, sp.17-CAY, sp.18-CAY, sp.1Guyafor, sp.2, sp.2-CAY, sp.21_DS, sp.22-CAY, sp.23-CAY, sp.24-CAY, sp.2BAF, sp.2Guyafor, sp.3, sp.3-CAY, sp.30-CAY, sp.33-CAY, sp.34-CAY, sp.35-CAY, sp.35Guyafor, sp.36-CAY, sp.38Guyafor, sp.3BAF, sp.3Guyafor, sp.4-CAY, sp.42-CAY, sp.5-CAY, sp.5Guyafor, sp.6-CAY, sp.6Guyafor, sp.7-CAY, sp.70-CAY, sp.71-CAY, sp.7Guyafor, sp.8-CAY, sp.9-CAY, sp.B1, sp.B10, sp.B2, sp.B35, sp.B5, sp.B8, sp.B9, sp.C, sp.D, sp.FG13-Holst, sp.FG14-Holst, sp.FG2-Holst, sp.FG21-Holst, sp.FG9-Holst, sp.P1, sp.P11, sp.P22, sp.P3, sp.P33, sp.P4, sp.P6, sp.P9

See the list below:

```{r species}
t <- select(taxo, species) %>%
  unique() %>%
  filter(grepl("^sp", species))
unique(taxo$species) %>%
  sort() %>%
  paste0(collapse = ", ")
```

```{r speciestab}
unid <- c("Indet.", "sp", "sp.", "Schaeff.")
subtaxo <- c(
  "guyanensis subsp. Guyanensis", "acuminatum_subsp.juniperinum",
  "cuspidata_subsp.dura", "guyanensis subsp. Guyanensis",
  "guyanensis_subsp.duckeana", "opacum_subsp.rabelianum",
  "panacoco_var.panacoco", "persistens_subsp.aurantiaca",
  "tomentosa_subsp.maroniensis", "venosa_subsp.amazonica",
  "vera_subsp.affinis"
)
undesc <- c(
  "sp.1", "sp.1-CAY", "sp.10-CAY", "sp.11-CAY", "sp.12-CAY",
  "sp.13-CAY", "sp.14-CAY", "sp.15-CAY", "sp.16-CAY",
  "sp.17-CAY", "sp.18-CAY", "sp.1Guyafor", "sp.2", "sp.2-CAY",
  "sp.21_DS", "sp.22-CAY", "sp.23-CAY", "sp.24-CAY", "sp.2BAF",
  "sp.2Guyafor", "sp.3", "sp.3-CAY", "sp.30-CAY", "sp.33-CAY",
  "sp.34-CAY", "sp.35-CAY", "sp.35Guyafor", "sp.36-CAY",
  "sp.38Guyafor", "sp.3BAF", "sp.3Guyafor", "sp.4-CAY",
  "sp.42-CAY", "sp.5-CAY", "sp.5Guyafor", "sp.6-CAY", "sp.6Guyafor",
  "sp.7-CAY", "sp.70-CAY", "sp.71-CAY", "sp.7Guyafor", "sp.8-CAY",
  "sp.9-CAY", "sp.B1", "sp.B10", "sp.B2", "sp.B35", "sp.B5", "sp.B8",
  "sp.B9", "sp.C", "sp.D", "sp.FG13-Holst", "sp.FG14-Holst",
  "sp.FG2-Holst", "sp.FG21-Holst", "sp.FG9-Holst", "sp.P1", "sp.P11",
  "sp.P22", "sp.P3", "sp.P33", "sp.P4", "sp.P6", "sp.P9"
)
taxo %>%
  select(species) %>%
  unique() %>%
  summarise(
    "Unidentified" = sum(species %in% unid,
      na.omit = TRUE
    ),
    "Sub-taxonomic" = sum(species %in% subtaxo,
      na.omit = TRUE
    ),
    "Undescribed" = sum(species %in% undesc,
      na.omit = TRUE
    )
  ) %>%
  gather("type", "number") %>%
  kable(caption = "Number of species by issue type.")
```
