```{r setup}
#| message: false
#| include: false
library(tidyverse)
library(knitr)
```

# Raw {.unnumbered}

First we explored raw species data content. This corresponds to **version 4**.

## Taxonomy preparation

We got different types of misstructured taxonomic informations:

-   Good: BAFOG, Iwokrama, Kabo, Malinau, Manare, Montagne Tortue, Paracou, Tene, Uppangala
-   Vernacular only - taxonomic information only in vernacular name: Jari, Moju, Peteco, Tapajos Km114, Tapajos Km67
-   Scientific only - taxonomic information only in scientific name: Ecosilva, Jenaro Herrera, Kibale, Lesong, Mbaiki, Misiones, Nelliyampathy, STREK, Sungai Lalang, Ulu Muda, Antimary APU2, Antimary APU3, Corinto, Embrapa Acre, Gola, Iracema, Mil, "Pad Limo 2 Barracos", "Pad Limo Chico Bocao", "Pad Limo Cumaru", "Pad Limo Jatoba", "Pad Limo Pocao", "Pad Limo STCP", "PAD Limoeiro Tabocal", Peixoto, Tirimbina
-   Scientific in species - scientifc name in the species column: Sao Nicolau

We smoothed all taxonomic information with five columns: scientific, vernacular, family, genus, species (`derived_data/raw_taxo_preped_vX.tsv`).

```{r taxo_prep}
#| message: false
#| warning: false
vernacular_only <- c("Moju", "Peteco", "Jari", "Tapajos Km114", "Tapajos Km67")
scientific_only <- c(
  "Kibale", "Lesong", "Mbaiki", "Ecosilva", "Jenaro Herrera",
  "Misiones", "Ulu Muda", "Sungai Lalang", "Nelliyampathy", "STREK",
  "Antimary APU2", "Antimary APU3", "Corinto", "Embrapa Acre", "Gola",
  "Iracema", "Mil", "Pad Limo 2 Barracos", "Pad Limo Chico Bocao",
  "Pad Limo Cumaru", "Pad Limo Jatoba", "Pad Limo Pocao", "Pad Limo STCP",
  "PAD Limoeiro Tabocal", "Peixoto", "Tirimbina"
)
scientific_in_species <- c("Sao Nicolau")
raw_file <- "raw_taxonomy_v4.tsv"
path <- file.path("data", "raw_data", raw_file)
taxo_easy <- read_tsv(path) %>%
  rename_all(tolower) %>%
  filter(!(site %in% c(
    vernacular_only,
    scientific_only,
    scientific_in_species
  )))
taxo_vernacular_only <- read_tsv(path) %>%
  rename_all(tolower) %>%
  filter(site %in% vernacular_only) %>%
  separate(vernname, c("vernname", "scientificname"), sep = "\\[") %>%
  separate(vernname, "vernname", sep = "\\/") %>%
  mutate(scientificname = gsub("]", "", scientificname)) %>%
  separate(scientificname, c("genus", "species"), remove = FALSE)
taxo_scientific_only <- read_tsv(path) %>%
  rename_all(tolower) %>%
  filter(site %in% scientific_only) %>%
  separate(scientificname, c("genus", "species"), remove = FALSE)
taxo_scientific_in_species <- read_tsv(path) %>%
  rename_all(tolower) %>%
  filter(site %in% scientific_in_species) %>%
  separate(species, c("a", "species")) %>%
  select(-a)
taxo <- bind_rows(
  taxo_easy, taxo_vernacular_only,
  taxo_scientific_only, taxo_scientific_in_species
) %>%
  rename(scientific = scientificname, vernacular = vernname) %>%
  separate(genus, "genus", sep = "[\\.]", remove = FALSE) %>%
  separate(species, "species", sep = "[_\\.]", remove = FALSE) %>%
  mutate_all(str_squish)
write_tsv(taxo, "data/derived_data/raw_taxo_preped_v4.tsv")
```

## All

We had data for 41 sites:

```{r stats}
taxo %>%
  group_by(site) %>%
  summarise(
    family = length(unique(na.omit(family))),
    genus = length(unique(na.omit(genus))),
    species = length(unique(paste(genus, species))),
    scientific = length(unique(na.omit(scientific))),
    vernacular = length(unique(na.omit(vernacular)))
  ) %>% 
  kable(caption = "Numbers per taxonomic levels for each site.")
```

In total, we thus add 2,440 species before cleaning:

```{r totals}
taxo %>%
  summarise(
    family = length(unique(na.omit(family))),
    genus = length(unique(na.omit(genus))),
    species = length(unique(paste(genus, species))),
    scientific = length(unique(na.omit(scientific))),
    vernacular = length(unique(na.omit(vernacular)))
  ) %>% 
  kable(
    format.args = list(big.mark = ","),
    caption = "Total numbers per taxonomic levels."
  )
```

Species name construction using genus and species should be preferred as the scientific name column increases diversity with only slight differences:

```{r diffs}
taxo %>%
  mutate(species = paste(genus, species)) %>%
  select(species, scientific) %>%
  unique() %>%
  group_by(species) %>%
  mutate(N = n()) %>%
  ungroup() %>%
  filter(species == "Paracroton pendulus") %>%
  kable(caption = "Example of multiple scientific names for one species genus pair.") # nolint
```

```{r diffs_fig}
#| message: false
#| fig-cap: Scientific names per genus species pairs.
taxo %>%
  mutate(species = paste(genus, species)) %>%
  select(species, scientific) %>%
  unique() %>%
  group_by(species) %>%
  summarise(N = n()) %>%
  filter(N > 1) %>%
  ggplot(aes(N)) +
  geom_histogram() +
  theme_bw() +
  scale_y_sqrt() +
  scale_x_sqrt() +
  xlab("Scientific names per genus species pairs") +
  ylab("")
```

## Family

Families includes the following dubious names: #N/A, +, Indet, Indet., muerto, ND, No determinado, Not in DB, unid., unidentified, Unidentified

See the list below:

```{r family}
fam_wrong <- c(
  "#N/A", "+", "muerto", "Indet.", "Not in DB", "unid.",
  "Unidentified", "unidentified", "ND", "No determinado", "Indet"
)
unique(filter(taxo, !(family %in% fam_wrong))$family) %>%
  sort() %>%
  paste0(collapse = ", ")
```

With the following counts:

```{r familytab}
taxo %>%
  select(family, genus, species) %>%
  unique() %>%
  filter(family %in% fam_wrong) %>%
  summarise(N = n()) %>%
  kable(caption = "Number of family issues")
```

## Genus

Genus list:

```{r genus}
gen_wrong <- c(
  "", "#N/A", "A", "Indet", "nd", "sp", "sp1", "sp5", "Indetermined",
  "KERUING", "MERANTI", "MERSAWA", "BALAU", "muerto", "Unidentified",
  "Nd", "nd10", "nd11", "nd14", "nd15", "nd16", "nd17", "nd2", "nd22",
  "nd23", "nd25", "nd26", "nd27", "nd28", "nd29", "nd33", "nd34",
  "nd4", "nd6", "nd7", "nd9", "N", "F", "cf", "No", "UN", "Un", "Spt",
  "Unident", "Unknow", "unknown", "Unknown", "ni", "n"
)
t <- taxo %>%
  filter(!(genus %in% gen_wrong)) %>%
  select(genus) %>%
  unique() %>%
  filter(str_length(genus) <= 3)
unique(filter(taxo, !(genus %in% gen_wrong))$genus) %>%
  sort() %>%
  paste0(collapse = ", ")
```

Genera includes the following dubious names: `{r} paste(gen_wrong, collapse = ",")`.

With the following counts:

```{r genustab}
taxo %>%
  select(family, genus, species) %>%
  unique() %>%
  filter(genus %in% gen_wrong) %>%
  summarise(N = n()) %>%
  kable(caption = "Number of genus issues")
```

## Species

Species list:

```{r species}
sp_wrong <- c(
  "Indet", "Indet.", "sp", "sp.", "Schaeff", "Schaeff.", "Mast", "Welw",
  "Stapf", "Hook", "Gilg", "Benth", "flacouirtiaceae", "Determinado",
  "BERMINYAK", "TIDAK", "KUNING", "spp", "", "1", "10", "11", "12", "13",
  "14", "15", "16", "17", "18", "19", "2", "20", "21", "22", "23", "25",
  "27", "28", "3", "4", "5", "6", "7", "8", "9", "a", "A", "sp1", "sp5",
  "L", "muerto", "JAP1001", "JAP1028", "JAP1036", "JAP1291", "JAP1398",
  "JAP1457", "JAP1627", "JAP1725", "JAP1799", "JAP1856", "JAP1873",
  "JAP1882", "JAP1891", "jariensis", "jauari", "javanica", "javanicum",
  "JCO355", "JCO407", "JCO416", "JCO578", "JCO597", "JCO670", "JCO696",
  "JCO710", "JCO736", "GFL1876", "GFL1997", "GFL2022", "GFL2211",
  "GFL2386", "GFL2411", "GFL2488", "GFL2606", "GFL2703", "GFL2772",
  "GFL2941", "GFL3093", "GFL3170", "GFL3182", "GFL3224", "GFL3284",
  "GFL3295", "GFL3481", "GFL3661", "GFL3686", "GFL6037", "GHP102",
  "JAP1000", "id", "indet", "Ridl", "sp13", "sp2", "sp3", "sp53",
  "species", "X", "XIV", "XVI", "spp", "n", "sp2", "a", "int",
  "L", "sur", "P", "4", "sp3", "id", "X", "2", "23", "1", "XVI",
  "19", "17", "5", "7", "22", "21", "3", "10", "11", "18", "16", "8",
  "XIV", "20", "14", "dao", "13", "6", "9", "25", "12", "28", "15",
  "27", "pin", "cf", "nd"
)
t <- taxo %>%
  filter(!(species %in% sp_wrong)) %>%
  select(species) %>%
  unique() %>%
  filter(str_length(species) <= 3)
unique(filter(taxo, !(species %in% sp_wrong))$species) %>%
  sort() %>%
  paste0(collapse = ", ")
```

Species includes the following dubious names: `{r} paste(sp_wrong, collapse = ",")`.

With the following counts:

```{r speciestab}
taxo %>%
  select(family, genus, species) %>%
  unique() %>%
  filter(species %in% sp_wrong) %>%
  summarise(N = n()) %>%
  kable(caption = "Number of species issues")
```
