```{r setup}
#| message: false
#| include: false
library(tidyverse)
library(knitr)
```

# Cleaning {.unnumbered}

Then we started by cleaning raw species data content. This corresponds to **version 2**.

## Names corrections

First we cleaned family, genus and species names using contingency table built on the information of dubious names from the previous chapter.

```{r namescor}
#| message: false
#| warning: false
family_tab <- data.frame(
  family_corrected = NA,
  family_raw = c(
    "#N/A", "+", "muerto",
    "Indet.", "Not in DB", "ND", "No determinado", "unid.", "unidentified"
  ),
  family_reason = c(
    rep("missing", 3),
    rep("unidentified", 6)
  )
)
write_tsv(family_tab, "data/derived_data/family_correction_table_v2.tsv")
genus_tab <- data.frame(
  genus_corrected = NA,
  genus_raw = c(
    " #N/A", "A",
    "Indet.", "nd", "sp", "sp1", "sp5", "Indetermined", "No",
    "KERUING", "MERANTI", "MERSAWA",
    "BALAU", "muerto",
    "Indet.Apocynaceae", "Indet.Arecaceae",
    "Indet.Bignoniaceae", "Indet.Burseraceae",
    "Indet.Capparaceae", "Indet.Chrysobalanaceae", "Indet.Clusiaceae",
    "Indet.Combretaceae", "Indet.Dichapetalaceae", "Indet.Euphorbiaceae",
    "Indet.Fabaceae", "Indet.FabaceaeCaesalpinioideae(mimosoid_clade)",
    "Indet.FabaceaePapilionoideae", "Indet.Humiriaceae", "Indet.Indet.",
    "Indet.Lamiaceae", "Indet.Lauraceae", "Indet.Lecythidaceae",
    "Indet.Malvaceae", "Indet.MalvaceaeBombacoideae",
    "Indet.Melastomataceae", "Indet.Moraceae", "Indet.Myrtaceae",
    "Indet.Nyctaginaceae", "Indet.Olacaceae", "Indet.Rubiaceae",
    "Indet.Salicaceae", "Indet.Sapindaceae", "Indet.Sapotaceae",
    "Indet.Urticaceae", "Indet.Violaceae", "Indet.Vochysiaceae",
    "Indet.Anacardiaceae", "Indet.Annonaceae"
  ),
  genus_reason = c(
    rep("unidentified", 2),
    rep("missing", 12),
    rep("family-only", 34)
  )
)
write_tsv(genus_tab, "data/derived_data/genus_correction_table_v2.tsv")
species_tab1 <- data.frame(
  species_corrected = NA,
  species_raw = c(
    "Indet.", "sp", "sp.", "Schaeff.", "spp",
    "Mast", "Welw", "Stapf", "Hook", "Gilg", "Benth", "flacouirtiaceae",
    "Determinado", "BERMINYAK", "TIDAK", "KUNING",
    "sp1", "sp5", "L", "muerto",
    "sp.1", "sp.1-CAY", "sp.10-CAY", "sp.11-CAY", "sp.12-CAY",
    "sp.13-CAY", "sp.14-CAY", "sp.15-CAY", "sp.16-CAY",
    "sp.17-CAY", "sp.18-CAY", "sp.1Guyafor", "sp.2", "sp.2-CAY",
    "sp.21_DS", "sp.22-CAY", "sp.23-CAY", "sp.24-CAY", "sp.2BAF",
    "sp.2Guyafor", "sp.3", "sp.3-CAY", "sp.30-CAY", "sp.33-CAY",
    "sp.34-CAY", "sp.35-CAY", "sp.35Guyafor", "sp.36-CAY",
    "sp.38Guyafor", "sp.3BAF", "sp.3Guyafor", "sp.4-CAY",
    "sp.42-CAY", "sp.5-CAY", "sp.5Guyafor", "sp.6-CAY", "sp.6Guyafor",
    "sp.7-CAY", "sp.70-CAY", "sp.71-CAY", "sp.7Guyafor", "sp.8-CAY",
    "sp.9-CAY", "sp.B1", "sp.B10", "sp.B2", "sp.B35", "sp.B5", "sp.B8",
    "sp.B9", "sp.C", "sp.D", "sp.FG13-Holst", "sp.FG14-Holst",
    "sp.FG2-Holst", "sp.FG21-Holst", "sp.FG9-Holst", "sp.P1", "sp.P11",
    "sp.P22", "sp.P3", "sp.P33", "sp.P4", "sp.P6", "sp.P9"
  ),
  species_reason = c(
    rep("unidentified", 20),
    rep("undescribed", 65)
  )
)
species_tab2 <- data.frame(
  species_corrected = c(
    "guyanensis", "acuminatum", "cuspidata", "guyanensis",
    "guyanensis", "opacum", "panacoco", "persistens",
    "tomentosa", "venosa", "vera"
  ),
  species_raw = c(
    "guyanensis subsp. Guyanensis", "acuminatum_subsp.juniperinum",
    "cuspidata_subsp.dura", "guyanensis subsp. Guyanensis",
    "guyanensis_subsp.duckeana", "opacum_subsp.rabelianum",
    "panacoco_var.panacoco", "persistens_subsp.aurantiaca",
    "tomentosa_subsp.maroniensis", "venosa_subsp.amazonica",
    "vera_subsp.affinis"
  ),
  species_reason = "sub-taxonomic "
)
bind_rows(species_tab1, species_tab2) %>%
  write_tsv("data/derived_data/species_correction_table_v2.tsv")
corrected <- read_tsv("data/derived_data/raw_taxo_preped_v2.tsv") %>%
  rename(
    scientific_raw = scientific, vernacular_raw = vernacular,
    family_raw = family, genus_raw = genus, species_raw = species
  ) %>%
  left_join(read_tsv("data/derived_data/family_correction_table_v2.tsv")) %>%
  mutate(family_corrected = ifelse(is.na(family_reason),
    family_raw, family_corrected
  )) %>%
  left_join(read_tsv("data/derived_data/genus_correction_table_v2.tsv")) %>%
  mutate(genus_corrected = ifelse(is.na(genus_reason),
    genus_raw, genus_corrected
  )) %>%
  left_join(read_tsv("data/derived_data/species_correction_table_v2.tsv")) %>%
  mutate(species_corrected = ifelse(is.na(species_reason),
    species_raw, species_corrected
  ))
write_tsv(corrected, "data/derived_data/corrected_taxo_v2.tsv")
```

```{r namecortab}
#| message: false
#| warning: false
read_tsv("data/derived_data/corrected_taxo_v2.tsv") %>%
  select(family_reason, genus_reason, species_reason) %>%
  gather("level", "reason") %>%
  na.omit() %>%
  group_by(level, reason) %>%
  summarise(corrections = n()) %>%
  kable(
    format.args = list(big.mark = ","),
    caption = "Number of corrections per level and reason."
  )
```

## World Flora Online (WFO)

We used `WorldFlora` and `fuzzyjoin` for synonymy and misspelling corrections. Beware, the R package is very memory consuming and I used batches of species names in the code. you might want to adapt it to the capacity of your computer. I also subsampled the backbones to the family list from the data to increase speed, but this might not be optimal.

```{r wfo}
#| message: false
#| warning: false
#| eval: false
library(WorldFlora)
library(fuzzyjoin)
batch_size <- 200
dir.create("data/derived_data/wfo")
families <- read_tsv("data/derived_data/corrected_taxo_v2.tsv") %>%
  select(family_corrected) %>%
  unique()
wfo <- read_tsv("data/raw_data/WFO_Backbone.zip")
wfo_sub <- wfo %>%
  filter(family %in% c(
    families$family_corrected,
    "Huaceae", " Zygophyllaceae", "Ctenolophonaceae",
    "Lepidobotryaceae", "Thomandersiaceae"
  ))
write_tsv(wfo_sub, "data/derived_data/wfo/backbone.tsv")
taxo <- read_tsv("data/derived_data/corrected_taxo_v2.tsv") %>%
  filter(!is.na(genus_corrected)) %>%
  mutate(species_corrected = ifelse(is.na(species_corrected),
    "", species_corrected
  )) %>%
  transmute(scientific = paste(genus_corrected, species_corrected)) %>%
  arrange(scientific) %>%
  unique()
n_batch <- ceiling(nrow(taxo) / batch_size)
taxo <- taxo %>%
  mutate(batch = rep(seq_len(n_batch), each = batch_size)[seq_len(n())])
write_tsv(taxo, "data/derived_data/wfo/taxo.tsv")
for (i in 1:n_batch) {
  rm(taxo, cleaned)
  gc()
  taxo <- read_tsv("data/derived_data/wfo/taxo.tsv") %>%
    filter(batch == i)
  cleaned <- WFO.match.fuzzyjoin(
    spec.data = taxo$scientific,
    WFO.file = "data/derived_data/wfo/backbone.tsv"
  )
  write_tsv(
    cleaned,
    paste0("data/derived_data/wfo/wfo_out_v2_part", i, ".tsv")
  )
}
unlink("data/derived_data/wfo/taxo.tsv")
unlink("data/derived_data/wfo/backbone.tsv")
cleaned <- read_tsv(list.files("data/derived_data/wfo", full.names = TRUE))
write_tsv(cleaned, "data/derived_data/wfo_out_v2.tsv")
unlink("data/derived_data/wfo", recursive = TRUE)
```

2,693 species matched a species from the World Flora Online (WFO) except 8 that can be cured manually:

```{r matches}
#| message: false
#| warning: false
wfo <- read_tsv("data/derived_data/wfo_out_v2.tsv")
wfo %>%
  group_by(spec.name) %>%
  summarise(matched = all(Matched)) %>%
  group_by(matched) %>%
  summarise(N = n(), names = ifelse(!matched, paste(spec.name), "")) %>%
  unique() %>%
  kable(
    format.args = list(big.mark = ","),
    caption = "Number of matched and unmatched species."
  )
```

Among matched species, most were exact (not fuzzy misspelled species) and with a single match in WFO, still 154 had multiple matches and 292 were fuzzy misspelled species, both to investigate:

```{r fuzzy_unique}
#| message: false
wfo %>%
  filter(Matched) %>%
  group_by(spec.name) %>%
  summarise(fuzzy = all(Fuzzy), unique = all(Unique)) %>%
  group_by(fuzzy, unique) %>%
  summarise(N = n()) %>%
  mutate(
    fuzzy = ifelse(fuzzy, "fuzzy", "exact"),
    unique = ifelse(unique, "unique", "multiple")
  ) %>%
  pivot_wider(names_from = "unique", values_from = "N") %>%
  kable(
    format.args = list(big.mark = ","),
    caption = "Number of fuzzy and unique species."
  )
```

## Manual curation

The 8 species that did not match WFO were manually cured by searching manually on WFO and on the web. Some were missed because lacking the family name in the WFO backbone subset, some do not matched species names, some used the subspecies names, and other were really misspelled.

```{r manual_curration}
manual_curration <- data.frame(
  scientific_corrected = c(
    "Alseodaphne ceratoxylon",
    "Alseodaphne havilandii",
    "Balanites wilsoniana",
    "Brosimum obovata",
    "Chrysophyllum pachycarpa",
    "Fernandoa adolfi",
    "Handroanthus campestris",
    "Palaquium cochlariflorum"
  ),
  family_cleaned = c(
    "Lauraceae",
    "Lauraceae",
    "Moraceae",
    "Zygophyllaceae",
    "Sapotaceae",
    "Bignoniaceae",
    "Bignoniaceae",
    "Sapotaceae"
  ),
  genus_cleaned = c(
    "Alseodaphne",
    "Alseodaphne",
    "Brosimum",
    "Balanites",
    "Chrysophyllum",
    "Fernandoa",
    "Handroanthus",
    "Palaquium"
  ),
  species_cleaned = c(
    NA,
    NA,
    "acutifolium",
    "obovata",
    "lucentifolium",
    "adolfi-friderici",
    NA,
    "cochleariifolium"
  ),
  cleaning_type = "manual curration",
  cleaning_comment = c(
    "unknown species on wfo",
    "unknown species on wfo",
    "Brosimum acutifolium subsp.",
    "missing family in wfo subset",
    "Chrysophyllum lucentifolium subsp. pachycarpum",
    "issue with the - in species name",
    "unknown species on wfo",
    "very misspelled"
  )
)
kable(manual_curration, caption = "Manually curated taxonomy that did not match WFO.") # nolint
```

## Misspelling (fuzzy)

By default `WorldFlora` limits fuzzy distance to 4, which seems more than acceptable and could be increased. Indeed one of the unmatched species, *Palaquium cochlariflorum* might be *Palaquium cochleariifolium* with a fuzzy distance of 6, but this can also be dealt with manually.

```{r fuzzy_dist}
wfo %>%
  filter(Matched, Fuzzy) %>%
  group_by(spec.name) %>%
  summarise(fuzzy_dist = unique(Fuzzy.dist)) %>%
  group_by(fuzzy_dist) %>%
  summarise(N = n()) %>%
  kable(caption = "Number of misspelled species by the number of misspelled letters.") # nolint
```

However, among the misspelled species with multiple matches, 4 have multiple newly accepted names:

```{r fuzzy_multiple}
wfo %>%
  filter(Matched, Fuzzy, !Unique, New.accepted) %>%
  select(spec.name, scientificName) %>%
  unique() %>%
  group_by(spec.name) %>%
  filter(n() > 1) %>%
  summarise(scientificName = paste0(scientificName, collapse = ", ")) %>%
  kable(caption = "Fuzzy match with multiple newly accepted names.")
```

## Synonymy (multiple matches)

Similarly, among the correctly spelled species with multiple matches, 6 have multiple newly accepted names:

```{r exact_multiple}
wfo %>%
  filter(Matched, !Fuzzy, !Unique, New.accepted) %>%
  select(spec.name, scientificName) %>%
  unique() %>%
  group_by(spec.name) %>%
  filter(n() > 1) %>%
  summarise(scientificName = paste0(scientificName, collapse = ", ")) %>%
  kable(caption = "Exact match with multiple newly accepted names.")
```

## Synonymy & misspelling corrections

> **⚠️ Multiple newly accepted names should be discussed.**

However, for the moment we will randomly select the first newly accepted name in this case. We will thus correct synonymy and misspelling by:

1.  Using closest correct spelling
2.  Accepting single exact matches
3.  Accepting single newly accepted names for multiple matches
4.  Accepting the first newly accepted names for multiple matches in case of multiples (to be changed later)
5.  Using manual curation for species not matched

```{r wfo_clean}
wfo <- read_tsv("data/derived_data/wfo_out_v2.tsv") %>%
  filter(Matched)
wfo_exact <- wfo %>%
  filter(Unique)
wfo_mult <- wfo %>%
  filter(!Unique) %>%
  group_by(spec.name) %>%
  filter(New.accepted) %>%
  sample_n(1)
wfo_clean <- bind_rows(wfo_exact, wfo_mult) %>%
  select(
    spec.name, family, genus, specificEpithet,
    scientificName
  ) %>%
  rename(
    scientific_corrected = spec.name,
    family_cleaned = family,
    genus_cleaned = genus,
    species_cleaned = specificEpithet,
    scientific_cleaned = scientificName
  ) %>%
  mutate(
    cleaning_type = "wfo automatic",
    cleaning_comment = NA
  ) %>%
  bind_rows(manual_curration)
taxo_cleaned <- read_tsv("data/derived_data/corrected_taxo_v2.tsv") %>%
  mutate(scientific_corrected = ifelse(!is.na(genus_corrected),
    paste(
      genus_corrected,
      ifelse(!is.na(species_corrected),
        species_corrected,
        ""
      )
    ), NA
  )) %>%
  left_join(wfo_clean)
write_tsv(taxo_cleaned, "outputs/taxonomy_v2.tsv")
```

```{r wfo_clean_stat}
#| message: false
clean <- read_tsv("outputs/taxonomy_v2.tsv")
fam_cor <- select(clean, family_corrected, family_cleaned) %>%
  unique() %>%
  na.omit() %>%
  filter(family_corrected != family_cleaned)
gen_cor <- select(clean, genus_corrected, genus_cleaned) %>%
  unique() %>%
  na.omit() %>%
  filter(genus_corrected != genus_cleaned)
sp_cor <- select(clean, species_corrected, species_cleaned) %>%
  unique() %>%
  na.omit() %>%
  filter(species_corrected != species_cleaned)
data.frame(
  level = c("family", "genus", "species"),
  N = c(nrow(fam_cor), nrow(gen_cor), nrow(sp_cor))
) %>%
  kable(caption = "Number of synonymy, misspelling, and update corrections per taxonomic level") # nolint
```

## **Taxonomic data**

All taxonomic data are saved in `outputs/axonomy_v2.tsv` with the following columns:

-   site: the site name
-   scientific\_\*: all columns with the scientific name
-   vernacular\_\*: all columns with the vernacular name
-   family\_\*: all columns with the family name
-   genus\_\*: all columns with the genus name
-   species\_\*: all columns with the species name
-   \*\_raw: all columns with the raw information as read in the inventories, see analyses
-   \*\_corrected: all columns with the manual corrections of dubious names, see analyses
-   \*\_cleaned: all columns with the synonymy and misspelling corrections with WFO and manual corrections, see analyses, this is the final taxonomic classification that should be used in further analyses
-   \*\_reason: the reason for manual corrections of dubious names, see analyses
-   cleaning_type: WFO automatic cleaning or using manual curation
-   cleaning_comment: the reason for manual curation
