```{r setup}
#| message: false
#| include: false
library(tidyverse)
library(WorldFlora)
library(fuzzyjoin)
library(knitr)
```

# Cleaning {.unnumbered}

Then we started by cleaning raw species data content. This corresponds to **version 1**.

## Names corrections

First we cleaned family, genus and species names using contingency table built on the information of dubious names from the previous chapter. We obtained **1,032** corrections in total.

```{r namescor}
#| eval: false
family_tab <- data.frame(
  family_corrected = NA,
  family = c("Indet.", "Not in DB", "unid.", "unidentified", "#N/A", "+"),
  family_reason = c(
    rep("missing", 2),
    rep("family-only", 4)
  )
)
write_tsv(family_tab, "outputs/family_correction_table.tsv")
genus_tab <- data.frame(
  genus_corrected = NA,
  genus = c(
    " #N/A", "Indet.", "Indet.Apocynaceae", "Indet.Arecaceae",
    "Indet.Bignoniaceae", "Indet.Burseraceae",
    "Indet.Capparaceae", "Indet.Chrysobalanaceae", "Indet.Clusiaceae",
    "Indet.Combretaceae", "Indet.Dichapetalaceae", "Indet.Euphorbiaceae",
    "Indet.Fabaceae", "Indet.FabaceaeCaesalpinioideae(mimosoid_clade)",
    "Indet.FabaceaePapilionoideae", "Indet.Humiriaceae", "Indet.Indet.",
    "Indet.Lamiaceae", "Indet.Lauraceae", "Indet.Lecythidaceae",
    "Indet.Malvaceae", "Indet.MalvaceaeBombacoideae",
    "Indet.Melastomataceae", "Indet.Moraceae", "Indet.Myrtaceae",
    "Indet.Nyctaginaceae", "Indet.Olacaceae", "Indet.Rubiaceae",
    "Indet.Salicaceae", "Indet.Sapindaceae", "Indet.Sapotaceae",
    "Indet.Urticaceae", "Indet.Violaceae", "Indet.Vochysiaceae"
  ),
  genus_reason = c(
    rep("unidentified", 1),
    rep("missing", 1),
    rep("family-only", 32)
  )
)
write_tsv(genus_tab, "outputs/genus_correction_table.tsv")


species_tab1 <- data.frame(
  species_corrected = NA,
  species = c(
    "Indet.", "sp", "sp.", "Schaeff.",
    "sp.1", "sp.1-CAY", "sp.10-CAY", "sp.11-CAY", "sp.12-CAY",
    "sp.13-CAY", "sp.14-CAY", "sp.15-CAY", "sp.16-CAY",
    "sp.17-CAY", "sp.18-CAY", "sp.1Guyafor", "sp.2", "sp.2-CAY",
    "sp.21_DS", "sp.22-CAY", "sp.23-CAY", "sp.24-CAY", "sp.2BAF",
    "sp.2Guyafor", "sp.3", "sp.3-CAY", "sp.30-CAY", "sp.33-CAY",
    "sp.34-CAY", "sp.35-CAY", "sp.35Guyafor", "sp.36-CAY",
    "sp.38Guyafor", "sp.3BAF", "sp.3Guyafor", "sp.4-CAY",
    "sp.42-CAY", "sp.5-CAY", "sp.5Guyafor", "sp.6-CAY", "sp.6Guyafor",
    "sp.7-CAY", "sp.70-CAY", "sp.71-CAY", "sp.7Guyafor", "sp.8-CAY",
    "sp.9-CAY", "sp.B1", "sp.B10", "sp.B2", "sp.B35", "sp.B5", "sp.B8",
    "sp.B9", "sp.C", "sp.D", "sp.FG13-Holst", "sp.FG14-Holst",
    "sp.FG2-Holst", "sp.FG21-Holst", "sp.FG9-Holst", "sp.P1", "sp.P11",
    "sp.P22", "sp.P3", "sp.P33", "sp.P4", "sp.P6", "sp.P9"
  ),
  species_reason = c(
    rep("unidentified", 4),
    rep("undescribed", 65)
  )
)
species_tab2 <- data.frame(
  species_corrected = c(
    "guyanensis", "acuminatum", "cuspidata", "guyanensis",
    "guyanensis", "opacum", "panacoco", "persistens",
    "tomentosa", "venosa", "vera"
  ),
  species = c(
    "guyanensis subsp. Guyanensis", "acuminatum_subsp.juniperinum",
    "cuspidata_subsp.dura", "guyanensis subsp. Guyanensis",
    "guyanensis_subsp.duckeana", "opacum_subsp.rabelianum",
    "panacoco_var.panacoco", "persistens_subsp.aurantiaca",
    "tomentosa_subsp.maroniensis", "venosa_subsp.amazonica",
    "vera_subsp.affinis"
  ),
  species_reason = "sub-taxonomic "
)
bind_rows(species_tab1, species_tab2) %>%
  write_tsv("outputs/species_correction_table.tsv")
cleaned <- read_tsv("data/raw_data/raw_taxonomy.tsv") %>%
  rename_all(tolower) %>%
  filter(!(site %in% c(
    "Kibale", "Lesong", "Moju", "Mbaiki", "Peteco",
    "Misiones", "Ulu Muda", "Sg Lalang"
  ))) %>%
  left_join(read_tsv("outputs/family_correction_table.tsv")) %>%
  mutate(family = ifelse(!is.na(family_reason), family_corrected, family)) %>%
  select(-family_corrected) %>%
  left_join(read_tsv("outputs/genus_correction_table.tsv")) %>%
  mutate(genus = ifelse(!is.na(genus_reason), genus_corrected, genus)) %>%
  select(-genus_corrected) %>%
  left_join(read_tsv("outputs/species_correction_table.tsv")) %>%
  mutate(species = ifelse(!is.na(species_reason),
    species_corrected, species
  )) %>%
  select(-species_corrected)
write_tsv(cleaned, "outputs/taxonomy_names_corrected.tsv")
```

```{r namecortab}
#| message: false
#| warning: false
read_tsv("outputs/taxonomy_names_corrected.tsv") %>%
  select(family_reason, genus_reason, species_reason) %>%
  gather("level", "reason") %>%
  na.omit() %>%
  group_by(level, reason) %>%
  summarise(corrections = n()) %>%
  kable(
    format.args = list(big.mark = ","),
    caption = "Number of corrections per level and reason."
  )
```

## Synonymy

```{r}
#| message: false
#| warning: false
#| eval: false
cleaned <- WFO.match.fuzzyjoin(
  spec.data = taxo3$scientificname,
  WFO.file = "data/raw_data/WFO_Backbone.zip"
)
write_tsv(cleaned, "data/derived_data/cleaned.tsv")
```

```{r}
#| message: false
cleaned <- read_tsv("data/derived_data/cleaned.tsv")
cleaned %>%
  select(spec.name, Matched, family, genus, specificEpithet) %>%
  kable()
```

Options for cleaning taxonomy:

-   **worldflora**
-   plantR
-   U.Taxonstand
-   taxlist
-   bdclean
