```{r setup}
#| message: false
#| include: false
library(tidyverse)
library(knitr)
```

# Cleaning {.unnumbered}

Then we started by cleaning raw species data content. This corresponds to **version 1**.

## Names corrections

First we cleaned family, genus and species names using contingency table built on the information of dubious names from the previous chapter. We obtained **1,032** corrections in total.

```{r namescor}
#| eval: false
family_tab <- data.frame(
  family_corrected = NA,
  family = c("Indet.", "Not in DB", "unid.", "unidentified", "#N/A", "+"),
  family_reason = c(
    rep("missing", 2),
    rep("family-only", 4)
  )
)
write_tsv(family_tab, "outputs/family_correction_table.tsv")
genus_tab <- data.frame(
  genus_corrected = NA,
  genus = c(
    " #N/A", "Indet.", "Indet.Apocynaceae", "Indet.Arecaceae",
    "Indet.Bignoniaceae", "Indet.Burseraceae",
    "Indet.Capparaceae", "Indet.Chrysobalanaceae", "Indet.Clusiaceae",
    "Indet.Combretaceae", "Indet.Dichapetalaceae", "Indet.Euphorbiaceae",
    "Indet.Fabaceae", "Indet.FabaceaeCaesalpinioideae(mimosoid_clade)",
    "Indet.FabaceaePapilionoideae", "Indet.Humiriaceae", "Indet.Indet.",
    "Indet.Lamiaceae", "Indet.Lauraceae", "Indet.Lecythidaceae",
    "Indet.Malvaceae", "Indet.MalvaceaeBombacoideae",
    "Indet.Melastomataceae", "Indet.Moraceae", "Indet.Myrtaceae",
    "Indet.Nyctaginaceae", "Indet.Olacaceae", "Indet.Rubiaceae",
    "Indet.Salicaceae", "Indet.Sapindaceae", "Indet.Sapotaceae",
    "Indet.Urticaceae", "Indet.Violaceae", "Indet.Vochysiaceae"
  ),
  genus_reason = c(
    rep("unidentified", 1),
    rep("missing", 1),
    rep("family-only", 32)
  )
)
write_tsv(genus_tab, "outputs/genus_correction_table.tsv")
species_tab1 <- data.frame(
  species_corrected = NA,
  species = c(
    "Indet.", "sp", "sp.", "Schaeff.",
    "sp.1", "sp.1-CAY", "sp.10-CAY", "sp.11-CAY", "sp.12-CAY",
    "sp.13-CAY", "sp.14-CAY", "sp.15-CAY", "sp.16-CAY",
    "sp.17-CAY", "sp.18-CAY", "sp.1Guyafor", "sp.2", "sp.2-CAY",
    "sp.21_DS", "sp.22-CAY", "sp.23-CAY", "sp.24-CAY", "sp.2BAF",
    "sp.2Guyafor", "sp.3", "sp.3-CAY", "sp.30-CAY", "sp.33-CAY",
    "sp.34-CAY", "sp.35-CAY", "sp.35Guyafor", "sp.36-CAY",
    "sp.38Guyafor", "sp.3BAF", "sp.3Guyafor", "sp.4-CAY",
    "sp.42-CAY", "sp.5-CAY", "sp.5Guyafor", "sp.6-CAY", "sp.6Guyafor",
    "sp.7-CAY", "sp.70-CAY", "sp.71-CAY", "sp.7Guyafor", "sp.8-CAY",
    "sp.9-CAY", "sp.B1", "sp.B10", "sp.B2", "sp.B35", "sp.B5", "sp.B8",
    "sp.B9", "sp.C", "sp.D", "sp.FG13-Holst", "sp.FG14-Holst",
    "sp.FG2-Holst", "sp.FG21-Holst", "sp.FG9-Holst", "sp.P1", "sp.P11",
    "sp.P22", "sp.P3", "sp.P33", "sp.P4", "sp.P6", "sp.P9"
  ),
  species_reason = c(
    rep("unidentified", 4),
    rep("undescribed", 65)
  )
)
species_tab2 <- data.frame(
  species_corrected = c(
    "guyanensis", "acuminatum", "cuspidata", "guyanensis",
    "guyanensis", "opacum", "panacoco", "persistens",
    "tomentosa", "venosa", "vera"
  ),
  species = c(
    "guyanensis subsp. Guyanensis", "acuminatum_subsp.juniperinum",
    "cuspidata_subsp.dura", "guyanensis subsp. Guyanensis",
    "guyanensis_subsp.duckeana", "opacum_subsp.rabelianum",
    "panacoco_var.panacoco", "persistens_subsp.aurantiaca",
    "tomentosa_subsp.maroniensis", "venosa_subsp.amazonica",
    "vera_subsp.affinis"
  ),
  species_reason = "sub-taxonomic "
)
bind_rows(species_tab1, species_tab2) %>%
  write_tsv("outputs/species_correction_table.tsv")
cleaned <- read_tsv("data/raw_data/raw_taxonomy.tsv") %>%
  rename_all(tolower) %>%
  filter(!(site %in% c(
    "Kibale", "Lesong", "Moju", "Mbaiki", "Peteco",
    "Misiones", "Ulu Muda", "Sg Lalang"
  ))) %>%
  left_join(read_tsv("outputs/family_correction_table.tsv")) %>%
  mutate(family = ifelse(!is.na(family_reason), family_corrected, family)) %>%
  select(-family_corrected) %>%
  left_join(read_tsv("outputs/genus_correction_table.tsv")) %>%
  mutate(genus = ifelse(!is.na(genus_reason), genus_corrected, genus)) %>%
  select(-genus_corrected) %>%
  left_join(read_tsv("outputs/species_correction_table.tsv")) %>%
  mutate(species = ifelse(!is.na(species_reason),
    species_corrected, species
  )) %>%
  select(-species_corrected)
write_tsv(cleaned, "data/derived_data/taxonomy_names_corrected.tsv")
```

```{r namecortab}
#| message: false
#| warning: false
read_tsv("data/derived_data/taxonomy_names_corrected.tsv") %>%
  select(family_reason, genus_reason, species_reason) %>%
  gather("level", "reason") %>%
  na.omit() %>%
  group_by(level, reason) %>%
  summarise(corrections = n()) %>%
  kable(
    format.args = list(big.mark = ","),
    caption = "Number of corrections per level and reason."
  )
```

## World Flora Online (WFO)

We used `WorldFlora` and `fuzzyjoin` for synonymy and misspelling corrections.

```{r wfo}
#| message: false
#| warning: false
#| eval: false
library(WorldFlora)
library(fuzzyjoin)
taxo <- read_tsv("data/derived_data/taxonomy_names_corrected.tsv") %>%
  filter(!is.na(species)) %>%
  transmute(scientificname = paste(genus, species)) %>%
  arrange(scientificname) %>%
  unique()
cleaned <- WFO.match.fuzzyjoin(
  spec.data = taxo$scientificname,
  WFO.file = "data/raw_data/WFO_Backbone.zip"
)
write_tsv(cleaned, "outputs/wfo_out.tsv")
```

1525 species matched a species from the World Flora Online (WFO) except 4:

```{r matches}
#| message: false
#| warning: false
wfo <- read_tsv("outputs/wfo_out.tsv")
wfo %>%
  group_by(spec.name) %>%
  summarise(matched = all(Matched)) %>%
  group_by(matched) %>%
  summarise(N = n(), names = ifelse(!matched, paste(spec.name), "")) %>%
  unique() %>%
  kable(
    format.args = list(big.mark = ","),
    caption = "Number of matched and unmatched species."
  )
```

Among matched species, most were exact (not fuzzy misspelled species) and with a single match in WFO, still 113 had multiple matches and 159 were fuzzy misspelled species, both to investigate:

```{r fuzzy_unique}
#| message: false
wfo %>%
  filter(Matched) %>%
  group_by(spec.name) %>%
  summarise(fuzzy = all(Fuzzy), unique = all(Unique)) %>%
  group_by(fuzzy, unique) %>%
  summarise(N = n()) %>%
  mutate(
    fuzzy = ifelse(fuzzy, "fuzzy", "exact"),
    unique = ifelse(unique, "unique", "multiple")
  ) %>%
  pivot_wider(names_from = "unique", values_from = "N") %>%
  kable(
    format.args = list(big.mark = ","),
    caption = "Number of fuzzy and unique species."
  )
```

## Misspelling (fuzzy)

By default `WorldFlora` limits fuzzy distance to 4, which seems more than acceptable and could be increased. Indeed one of the unmatched species, *Palaquium cochlariflorum* might be *Palaquium cochleariifolium* with a fuzzy distance of 6, but this can also be dealt with manually.

```{r fuzzy_dist}
wfo %>%
  filter(Matched, Fuzzy) %>%
  group_by(spec.name) %>%
  summarise(fuzzy_dist = unique(Fuzzy.dist)) %>%
  group_by(fuzzy_dist) %>%
  summarise(N = n()) %>%
  kable(caption = "Number of misspelled species by the number of misspelled letters.") # nolint
```

However, among the misspelled species with multiple matches, 4 have multiple newly accepted names:

```{r fuzzy_multiple}
wfo %>%
  filter(Matched, Fuzzy, !Unique, New.accepted) %>%
  select(spec.name, scientificName) %>%
  unique() %>%
  group_by(spec.name) %>%
  filter(n() > 1) %>%
  summarise(scientificName = paste0(scientificName, collapse = ", ")) %>%
  kable(caption = "Fuzzy match with multiple newly accepted names.")
```

## Synonymy (multiple matches)

Similarly, among the correctly spelled species with multiple matches, 6 have multiple newly accepted names:

```{r exact_multiple}
wfo %>%
  filter(Matched, !Fuzzy, !Unique, New.accepted) %>%
  select(spec.name, scientificName) %>%
  unique() %>%
  group_by(spec.name) %>%
  filter(n() > 1) %>%
  summarise(scientificName = paste0(scientificName, collapse = ", ")) %>%
  kable(caption = "Exact match with multiple newly accepted names.")
```

## Synonymy & misspelling corrections

> **⚠️ Multiple newly accepted names should be discussed.**

However, for the moment we will randomly select the first newly accepted name in this case. We will thus correct synonymy and misspelling by:

1.  Using closest correct spelling
2.  Accepting single exact matches
3.  Accepting single newly accepted names for multiple matches
4.  Accepting the first newly accepted names for multiple matches in case of multiples (to be changed later)

```{r wfo_clean}
#| eval: false
taxo <- read_tsv("data/derived_data/taxonomy_names_corrected.tsv") %>%
  mutate(scientificname = paste(genus, species)) %>%
  arrange(site, scientificname) %>%
  unique()
wfo <- read_tsv("outputs/wfo_out.tsv") %>%
  filter(Matched)
wfo_exact <- wfo %>%
  filter(Unique)
wfo_mult <- wfo %>%
  filter(!Unique) %>%
  group_by(spec.name) %>%
  filter(New.accepted) %>%
  sample_n(1)
wfo_clean <- bind_rows(wfo_exact, wfo_mult) %>%
  select(
    spec.name, family, genus, specificEpithet,
    scientificName
  ) %>%
  rename(
    scientificname = spec.name,
    family_wfo = family,
    genus_wfo = genus,
    species_wfo = specificEpithet,
    scientificname_wfo = scientificName
  )
write_tsv(wfo_clean, "outputs/wfo_correction_table.tsv")
left_join(taxo, wfo_clean) %>%
  write_tsv("data/derived_data/taxonomy_names_cleaned.tsv")
```

```{r wfo_clean_stat}
#| message: false
clean <- read_tsv("data/derived_data/taxonomy_names_cleaned.tsv")
fam_cor <- select(clean, family, family_wfo) %>%
  unique() %>%
  na.omit() %>%
  filter(family != family_wfo)
gen_cor <- select(clean, genus, genus_wfo) %>%
  unique() %>%
  na.omit() %>%
  filter(genus != genus_wfo)
sp_cor <- select(clean, species, species_wfo) %>%
  unique() %>%
  na.omit() %>%
  filter(species != species_wfo)
data.frame(
  level = c("family", "genus", "species"),
  N = c(nrow(fam_cor), nrow(gen_cor), nrow(sp_cor))
) %>%
  kable(caption = "Number of synonymy, misspelling, and update corrections per taxonomic level") # nolint
```
