```{r setup}
#| message: false
#| include: false
library(tidyverse)
library(ape)
library(phylobase)
library(ggtree)
```

# Phylogeny {.unnumbered}

We used cleaned taxonomy to build a global phylogeny using `V.PhyloMaker`. This corresponds to **version 3**.

```{r phylo}
#| eval: false
library(V.PhyloMaker)
library(ape)
splist <- read_tsv("outputs/taxonomy_v3.tsv") %>%
  select(species_cleaned, genus_cleaned, family_cleaned) %>%
  na.omit() %>%
  unique() %>%
  rename_all(~ gsub("_cleaned", "", .)) %>%
  mutate(species = paste(genus, species))
tree <- phylo.maker(
  sp.list = splist,
  tree = GBOTB.extended,
  nodes = nodes.info.1,
  scenarios = "S3"
)
write.tree(tree$scenario.3, "outputs/phylogeny_v3.tree")
```

We can see the results with all families (*a better representation might be found*):

```{r phyloall}
#| fig-width: 12
#| fig-height: 6
#| message: false
#| warning: false
#| fig-cap: "Global phylogeny of species across all sites."
splist <- read_tsv("outputs/taxonomy_v3.tsv") %>%
  select(species_cleaned, genus_cleaned, family_cleaned) %>%
  na.omit() %>%
  unique() %>%
  rename_all(~ gsub("_cleaned", "", .)) %>%
  mutate(species = paste(genus, species))
phylo <- ape::read.tree("outputs/phylogeny_v3.tree")
phylo_df <- phylo4d(
  phylo,
  data.frame(species = gsub("_", " ", phylo$tip.label)) %>%
    left_join(splist)
)
ggtree(phylo_df, aes(col = family), layout = "circular") +
  theme_tree(legend.text = element_text(face = "italic", size = 6)) +
  scale_color_discrete("")
```

And we can further explore the phylogeny per plot compared to the global phylogeny:

```{r phylosite}
#| fig-width: 12
#| fig-height: 12
#| message: false
#| warning: false
#| fig-cap: "Global phylogeny of species across all sites with taxa coloured per site." #nolint
phylo_site <- function(sit) {
  taxo <- read_tsv("outputs/taxonomy_v3.tsv") %>%
    filter(site == sit) %>%
    select(species_cleaned, genus_cleaned, family_cleaned) %>%
    na.omit() %>%
    unique() %>%
    rename_all(~ gsub("_cleaned", "", .)) %>%
    mutate(species = paste(genus, species))
  phylo_df@data$site <- phylo_df@data$species %in% taxo$species
  ggtree(phylo_df, aes(col = site, alpha = site), layout = "circular") +
    scale_color_manual(
      guide = "none",
      values = c("white", "darkgreen", "white")
    ) +
    scale_alpha_manual(
      guide = "none",
      values = c(0, 1, 0)
    ) +
    ggtitle(sit)
}
sites <- unique(read_tsv("outputs/taxonomy_v3.tsv")$site)
figs <- lapply(sites, phylo_site)
cowplot::plot_grid(plotlist = figs)
```

## **Phylogenic data**

The full phylogeny is saved in `outputs/phylogeny_vX.tree` in parenthetic format using the Newick (also known as New Hampshire) format.
